---
title: "Epilepsy Analysis"
output: pdf_document
date: "2024-11-20"
---

# Statistical Analysis of Epilepsy Dataset

### Libraries Used

```{r include=FALSE}
library(dplyr)
library(ggplot2)
library(caret)
library(pscl)
library(tidyverse)
library(broom)
library(tidycmprsk)
library(ggpubr)
library(tinytex)
```

```{r}
df <- read.csv('epilepsy.csv') # Read in dataset

df <- na.omit(df) # Remove all empty values

head(df) # Peak
```

```{r}
summary(df)
```

# Logistic Regression Model

Fit Logit Model

```{r}
model <- glm(Rem12~treat, family="binomial", data=df)
options(scipen=999)
summary(model)
```

N.B: The intercept is the CBZ group

```{r}
exp(coef(model)) # Odds Ratio
```

```{r}
exp(confint(model)) # Confidence Intervals of the models
```

```{r echo=FALSE}
tbl_regression(model, exponentiate = TRUE, label = list(treat ~ "Treatment Type")) 
```

## Visualisation

1.  Bar plot for predicted probabibilities

```{r echo=FALSE}
library(ggplot2)

# Generate probabilities for treatment levels
treatment_levels <- data.frame(treat = c("CBZ", "LTG"))
treatment_levels$predicted_prob <- predict(model, newdata = treatment_levels, type = "response")

treatment_levels$predicted_prob <- (treatment_levels$predicted_prob * 100)

# Plot the probabilities
ggplot(treatment_levels, aes(x = treat, y = predicted_prob, fill = treat)) +
  geom_col() +
  labs(title = "Predicted Probabilities by Treatment",
       x = "Treatment",
       y = "Predicted Probability (%)") +
  theme_pubr() +
  theme(
    legend.position = "None"
  ) +
  scale_fill_manual(values = c("CBZ"="#6D8325FF", "LTG"="#4A3587FF"))  +
  scale_y_continuous(breaks = seq(0, 100 , by = 5)) 

```

2.  ROC Curve

```{r echo=FALSE}
library(pROC)
predictions_test <- predict(model, newdata = df, type = "response")
roc_curve <- roc(df$Rem12, predictions_test)
ggroc(roc_curve) +
  theme_pubr() 
auc(roc_curve)  # Area Under Curve
```

3.  Forest Plot

```{r echo=FALSE}
library(forestplot)
dfmodel <- tidy(model, conf.int = TRUE)
Drug <- c("CBZ", "LTG")

dfmodel1 <- cbind(Drug, dfmodel)
forestplot(labeltext =dfmodel1$Drug ,
           mean = dfmodel1$estimate,
        lower = dfmodel1$conf.low,
        upper = dfmodel1$conf.high,
        xlab = "Adjusted Coefficients and 95% Confidence Intervals",
        boxsize = 0.1,
        col = fpColors(box = "black", line = "black", summary = "black",
                                     zero = "#4A3587FF"),
        txt_gp = fpTxtGp(label  = gpar (cex = 1.0),
                         xlab = gpar(cex = 1.0),
                         ticks = gpar (cex = 1.0),
                         title = gpar (cex = 1.0)),
        grid = TRUE,
        title = "Estimating the Effects of Treatment on \n Remission")
```

# Gender Interaction Model

This model assess the interaction between patient gender and treatment group.

```{r include=FALSE}
modelII <- glm(Rem12 ~ treat * sex, data = df, family = binomial)
options(scipen=999)
summary(modelII)
exp(confint(modelII))
```

```{r include=FALSE}
or <- exp(coef(modelII)) # Odds Ratio
or
```

```{r include=FALSE}
odds_df <- data.frame(
  Variable = names(or),
  OddsRatio = or
) %>%
  mutate(
    Variable = case_when(
      Variable == "(Intercept)" ~ "CBZ, Female (Intercept)",
      Variable == "treatLTG" ~ "LTG",
      Variable == "sexM" ~ "Male",
      Variable == "treatLTG:sexM" ~ "LTG, Male"
    )
  )

odds_df
```

```{r echo=FALSE}
tbl_regression(modelII, exponentiate = TRUE, label = list(treat ~ "Treatment Type")) 
```

1.  Bar Plot of Odds Ratios

```{r echo=FALSE}
ggplot(odds_df, aes(x = reorder(Variable, OddsRatio), y = OddsRatio)) +
  geom_bar(stat = "identity", fill = "skyblue") +  # Bar plot with skyblue color
  coord_flip() +  # Flip coordinates to make it horizontal
  labs(
    title = "Odds Ratios from \nLogistic Regression Model",
    x = "Variable",
    y = "Odds Ratio"
  ) +
  theme_pubr()  # Clean theme
```

2.  Forest Plot

```{r echo=FALSE}
dfmodelx <- tidy(modelII, conf.int = TRUE)
Term <- c("CBZ, Female (Intercept)", "LTG", "Male", "LTG, Male")

dfmodelx <- cbind(dfmodelx, Term)
forestplot(labeltext =dfmodelx$Term ,
           mean = dfmodelx$estimate,
        lower = dfmodelx$conf.low,
        upper = dfmodelx$conf.high,
        xlab = "Adjusted Coefficients and 95% Confidence \nIntervals",
        boxsize = 0.1,
        col = fpColors(box = "black", line = "black", summary = "black",
                                     zero = "steelblue"),
        txt_gp = fpTxtGp(label  = gpar (cex = 1.0),
                         xlab = gpar(cex = 1.0),
                         ticks = gpar (cex = 1.0),
                         title = gpar (cex = 1.0)),
        grid = TRUE,
        title = "Estimating the Effects of Treatment on \n Remission")
```

```{r}
drop1(modelII, .~., test = "Chisq")
```

## Making predictions with our model

```{r include=FALSE}
new_data <- expand.grid(
  treat = c("CBZ", "LTG"),
  sex = c("M", "F")
)
new_data$predicted_prob <- predict(modelII, newdata = new_data, type = "response")
new_data$predicted_prob <- (new_data$predicted_prob * 100)
```

```{r echo=FALSE}
# Plot predicted probabilities
ggplot(new_data, aes(x = treat, y = predicted_prob, fill = sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Predicted Probabilities by Treatment and Gender",
       x = "Treatment", y = "Predicted Probability") +
  scale_fill_manual(values = c("#6D8325FF", "#4A3587FF")) +
  theme_pubr() +
  labs(fill = "Sex")

```

# Time to Event Analysis

```{r include=FALSE}
library(survival)
library(survminer)
library(joineR)

dfII <- df # Create a copy of the main dataframe

dfII$treat <- factor(dfII$treat) # Make treatment group a factor
```

```{r}
# Make categorical variables factors
dfII$wdlcode <- factor(dfII$wdlcode, levels = c(0, 1, 2), 
                           labels = c("NOT", "UAE", "ISC")) 
dfII$treat <- factor(dfII$treat)
dfII$id <- factor(dfII$id)
```

```{r echo=FALSE}
hist(dfII$withtime)
```

```{r}
survobj <- Surv(time = dfII$withtime, event = dfII$censall) # Define survival objects with time variable and event of interest

head(survobj)
```

```{r}
# Fit Kaplan-Meier Survival Curves for Treatment Groups
km_fit <- survfit(Surv(withtime, censall) ~ treat, data = dfII)
summary(km_fit)
```

Survival Plots by Treatment Groups

```{r echo=FALSE}
library(ggfortify)

ggsurvplot(km_fit, data = dfII, conf.int = TRUE, pval = TRUE, risk.table = TRUE)
```

## Log Rank Test between Groups

```{r}
survdiff(Surv(withtime, censall) ~ treat, data = df)
```

## Cox Prop Hazards Model

```{r}
survb <- Surv(df$withtime, df$censall)

cox_model <- coxph(survb ~ treat, data =df)
summary(cox_model)

ci <- exp(confint(cox_model))
```

```{r}
ci # Confidence Intervals
```

Forest Plot

```{r echo=FALSE}
ggforest(cox_model, data = df)
```

### Assess reasonableness of underlying cox model assumptions

```{r}
# Proportional Hazards 
ph_assumption <- cox.zph(cox_model)

ph_assumption
```

```{r}
library(ggpubr)
# Linearity
plot(ph_assumption,
     xlab = "Time to Withdrawal (days)",                   # Change x-axis title
     ylab = "Time Varying Effect of Treatment (Beta(t))") 


# Add a horizontal reference line at y = 0
abline(h = 0, col = "red", lty = 2, lwd = 2)

mtext("p-value = 0.28", side = 3, line = 2)

legend(
  "bottomright",                                 # Position
  legend = c(
    "Smoothed Treatment Effect", 
    "95% Confidence Interval", 
    "Horizontal Reference (y = 0)"
  ),                                             # Legend text
  col = c("black", "black", "red"),             # Colors for each line
  lty = c(1, 2, 2),                             # Line types
  lwd = c(2, 1, 2), 
  xpd =TRUE, 
  cex = 0.4,
  inset = c(0.2, 0.2)  # Adjust legend text size
)

```

```{r eval=FALSE, include=FALSE}
library(cowplot)
ggcoxzph(ph_assumption, se = TRUE, resid = TRUE) 
```

```{r}
# Assessing the Overall Model Fit

# Likelihood ratio test
anova(cox_model, test = "Chisq")

#
AIC(cox_model)
```

# Competing Risk Analysis

```{r}
library(cmprsk)
library(survival)
library(crrstep)
```

```{r}
# Cummulative Incidence Function

dfIIx <- df



dfIIx$time <- dfIIx$withtime
dfIIx$group <- dfIIx$treat

dfIIx$status <- factor(dfIIx$wdlcode, labels = c("CEN", "UAE", "ISC"), levels = 0:2) # These are the three competing outcomes of interest

dfIIx <- dfIIx[dfIIx$status != "CEN", ]
## this is used to estimate the probability of each event over time
```

```{r}
cif <- cuminc(ftime = dfIIx$time,
              fstatus = dfIIx$status,
              group=dfIIx$group)

```

```{r}
cif
```

```{r}
names(cif)
groups <- c("CBZ UAE", "LTG UAE", "CBZ ISC" ,"LTG ISC")
cif_list <- list()
```

```{r}
for (group in groups) {
  cif_list[[group]] <- data.frame(
    time = cif[[group]]$time,
    estimate = cif[[group]]$est,
    group = group
  )
}

cif_data <- bind_rows(cif_list)
```

```{r}
ggplot(cif_data, aes(x = time, y = estimate, color = group, linetype = group)) +
  geom_line() +
  labs(
    x = "Time to Withdrawal (days)", 
    y = "Cumulative Incidence"
  ) +
  theme_pubr() +
  theme(legend.title = element_blank()) +
  scale_color_manual(values = c("#6D8325FF","#6D8325FF","#4A3587FF", "#4A3587FF")) +
  annotate("text", x = 1500, y = 0.20, label = "UAE: p = 0.0042", color = "#6D8325FF", size = 3, fontface = "bold") +
  annotate("text", x = 1500, y = 0.15, label = "ISC: p = 0.5315", color="#4A3587FF", size = 3) 
```

## Gray's Test

```{r}
# Gray test assesses whether CIF is significantly different between groups for each event.

gt <- cuminc(ftime = dfIIx$time,
              fstatus = dfIIx$status,
              group=dfIIx$group)
gt
```

## Fine and Gray Test

```{r include=FALSE}
fgdf <- dfIIx

#fgdf$treat <- factor(fgdf$treat, levels = c("CBZ", "LTG"))

fgdf$wdlcode <- as.numeric(fgdf$wdlcode)
str(fgdf)
```

```{r}
fgdf$etime <- fgdf$time

fgdf$event <- fgdf$wdlcode

fgdf$event2 <- factor(fgdf$event, 0:2, labels = c("CEN", "UAE", "ISC"))
```

```{r}
table(fgdf$event2)
```

```{r}
isc <- finegray(Surv(etime, event2) ~ treat, data=fgdf, etype="ISC")
```

```{r}
uae <- finegray(Surv(etime, event2) ~ treat, data=fgdf, etype="UAE")
```

```{r}
fgfitisc <- coxph(Surv(fgstart, fgstop, fgstatus) ~ treat, data=isc,
weight= fgwt)

fgfituae <- coxph(Surv(fgstart, fgstop, fgstatus) ~ treat, data=uae,
weight= fgwt)
```

```{r}
summary(fgfitisc)
```

```{r}
summary(fgfituae)
```

# Findings

1.  Lamotrigine demonstrated a slight advantage over Carbamazepine in helping participants stay seizure-free for 12 months, though this difference was not statistically significant.
2.  The gender-specific analysis showed women on CBZ had the lowest overall probability of achieving 12-month remission. Overall, no significant differences were observed between men and women in their response to treatment
3.  Survival analysis presented a clear comparison between the AEDs, with the Lamotrigine group consistently having lower withdrawal rates than Carbamazepine over the study period. LTG also had the lowest withdrawal rates due to unexpected adverse events. Lamotrigine may have been a more tolerable treatment and may have a reduced risk of withdrawal, ultimately making it a preferable option to improve adherence and tolerability.
4.  The competing risks analysis also highlights notable differences in treatment effects based on withdrawal causes. In withdrawal due to UAE, Lamotrigine had a statistically significant lower cumulative incidence suggesting it may be preferable due to fewer adverse effects.

# Limitations

1.  Models were not fitted with other variables in mind i.e Age & Seizure Frequency.
